### 2025-10-05 thesis/design/index.tex
Further improve the design section
Split and finish the sampling section
Write more of the indexing section

The core of the proposed design is the query function $nq$ which utilizes
example selection ($\sigma$), query preprocessing, schema subsetting ($\phi$),
model inference ($\pi$), self-correction ($\rho$) and majority voting ($\nu$)
in order to convert a natural language query $q$ into a corresponding SQL query $\omega$.

@@

Prior to using $nq$ to answer natural language questions, historical datasets must be sampled
such that an embedding space $\mathcal{V}$ and distance index $\mathcal{D}$ can be supplied
to $nq$.

\subimport{}{sampling}

>>>>

The core of the proposed design is the query function $nq$ which utilizes
example selection ($\sigma$), schema subsetting ($\phi$), model inference ($\pi$),
self-correction ($\rho$) and majority voting ($\nu$) in order to convert a natural
language query $q$ into a corresponding SQL query $\omega$.

@@

Prior to using $nq$ to answer natural language questions, historical datasets must be sampled
and indexed such that an embedding space $v$ and distance index $d$ can be supplied to $nq$.

\subimport{}{sampling}
\subimport{}{indexing}

### 2025-10-05 thesis/design/indexing.tex
Write more of the indexing section

For example $\omega_{ground}$ the question ``Give me all contacts for the user with the id 10''
might look very different depending on the database schema at hand, thus only taking into
account the similarity of the natural language question will yield inferior results.

\vspace{1em}

\begin{figure}[ht]
\hfill
\begin{minipage}[b]{0.45\linewidth}
\centering
\begin{verbatim}
CREATE TABLE users (
  id TEXT PRiMARY KEY
);

CREATE TABLE contacts (
  user_id TEXT NOT NULL,
  name TEXT NOT NULL,
  FOREIGN KEY (user_id) ON users(id),
  PRIMARY KEY (user_id, name)
);
\end{verbatim}
\caption{Normalized schema}
\end{minipage}
\hfill
\begin{minipage}[b]{0.35\linewidth}
\centering
\begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY,
  contacts TEXT[]
);
\end{verbatim}
\caption{Denormalized schema}
\end{minipage}
\hfill
\end{figure}

\vspace{1em}

>>>>

For example $\omega_{ground}$ the question ``Give me all contacts for the user with the id 10''
might look different depending on the database schema at hand, thus only selecting
samples based on the similarity of the natural language question will yield inferior
sample quality.

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY
);

CREATE TABLE contacts (
  user_id TEXT NOT NULL,
  name TEXT NOT NULL,
  FOREIGN KEY (user_id) ON users(id),
  PRIMARY KEY (user_id, name)
);
    \end{verbatim}
    \caption{Normalized schema}
    \label{figure:schema-normalized}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.35\linewidth}
    \begin{verbatim}
CREATE TABLE users (
  id TEXT PRIMARY KEY,
  contacts TEXT[] NOT NULL
);
    \end{verbatim}
    \vspace{3.75em}
    \caption{Denormalized schema}
    \label{figure:schema-denormalized}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

Given the database schemas in figures~\ref{figure:schema-normalized} and~\ref{figure:schema-denormalized}
respective definitions of $\omega_{ground}$ would be:

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \begin{verbatim}
SELECT contacts.name
FROM users
JOIN contacts
  ON contacts.user_id = users.id
WHERE users.id = 10;
    \end{verbatim}
    \caption{SQL JOIN selection}
    \label{figure:sql-join-example}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.35\linewidth}
    \centering
    \begin{verbatim}
SELECT contacts
FROM users
WHERE id = 10;
    \end{verbatim}
    \vspace{1.3em}
    \caption{SQL Array selection}
    \label{figure:sql-array-example}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

As shown in the figures~\ref{figure:sql-join-example} and~\ref{figure:sql-array-example}
the structural similarity of the underlying database schema is a crucial component
of the relevance of a example.

\subsubsection{Graph Representation}

In order to determine structural similarity of database schemas we propose using
graph representation of database schemas as a data definition language (DDL)
independent representation of the databases structure. Furthermore choosing a graph
representation allows us to use prevalent methods for measuring graph similarity
(eg, \textcolor{red}{Weisfeiler-Lehman graph kernels, cite}).

We define the structural similarity of two databases as the toplogical distance
of the respective \textsc{Natural}-graphs. The \textsc{Natural}-graph representation
only models tables, columns, relationships, datatypes and constraints, omitting 
table names, column names and domain terminology.

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
      \node[main] (1) {$t_1$}; 
      \node[main] (2) [below of=1] {$c_1$}; 
      \node[main] (3) [right of=2] {$t_2$}; 
      \node[main] (4) [below left of=3] {$c_3$}; 
      \node[main] (5) [below right of=3] {$c_4$};

      \draw (1) -- (2); 
      \draw (2) -- (4); 
      \draw (3) -- (4); 
      \draw (3) -- (5); 
      \draw (1) to [out=0, in=90, looseness=1] (3); 
    \end{tikzpicture} 
    \caption{Normalized graph repr.}
    \label{figure:normalized-graph-representation}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
      \node[main] (1) {$t_1$}; 
      \node[main] (2) [below of=1] {$c_1$}; 
      \node[main] (3) [below right of=1] {$c_2$};

      \draw (1) -- (2); 
      \draw (1) -- (3); 
    \end{tikzpicture} 
    \caption{Denormalized graph repr.}
    \label{figure:denormalized-graph-representation}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

Now measuring the distance between the graphs displayed in~\ref{figure:normalized-graph-representation} and~\ref{figure:denormalized-graph-representation}

### 2025-10-05 thesis/design/sampling.tex
Write more of the indexing section

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes an
embedding space to both traverse the masked natural language query $q'$
and the respective masked SQL query $\omega'$. This allows $\sigma$ to use
cosine distance to search through historically observed question answer pairs;
eg, from previous user interactions or prevalent training datasets like
\textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.

>>>>

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
cosine similarity in an embedding space $v$ to search for examples based on the
masked natural language query $q'$ and the respective masked SQL query $\omega'$.
This allows $\sigma$ to use cosine distance to search through historically observed
question answer pairs; eg, from previous user interactions or prevalent training
datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.
